Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-09-03 Manuscript/manuscript.Rmd
grammar

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed important 3-dimensional (3D) spatial constructs within the human genome such as loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. Among these, TADs represent regions on the linear genome, at megabase scale, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics like gene regulation and cell differentiation, an example being the mechanism by which enhancers loop to come into contact with distal promoters to initiate transcription within these domains [@Lupianez:2015aa; @Franke:2016aa]. Likewise, disruption of TAD boundaries has been shown to lead to developmental disorders [@lupianez2016breaking], such as cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@Sun:2018aa; @Meaburn:2007aa]. Therefore, identifying the location of TAD boundaries remains a top priority in our goal to understand the functionality of the human genome.

@@

Several methods have been developed to predict TAD boundary location using genomic annotations. However, most ignore key characteristics of 3D genomic data that are detrimental to model performance. A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, such high level of resolution is likely to introduce heavily imbalanced classes created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. *HiCFeat* did not the address class imbalance. Instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations, resulting in a zero inflated model. <!--This, likely being further exacerbated by the class imbalance.--> Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other re-sampling solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in base pairs between ChIP-seq annotations and genomic bins in TAD boundary prediction, which offer a more spatial measure of association.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed important 3-dimensional (3D) spatial constructs within the human genome such as loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. Among these, TADs represent regions on the linear genome, at megabase scale, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics like gene regulation and cell differentiation, an example being the mechanism by which enhancers loop to come into contact with distal promoters to initiate transcription within these domains [@Lupianez:2015aa; @Franke:2016aa]. Likewise, disruption of TAD boundaries has been shown to lead to developmental disorders [@lupianez2016breaking], cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@Sun:2018aa; @Meaburn:2007aa]. Therefore, identifying the location of TAD boundaries remains a top priority in our goal to understand the functionality of the human genome.

@@

Several methods have been developed to predict TAD boundary location using genomic annotations. However, most ignore key characteristics of 3D genomic data that are detrimental to both model performance and accurate boundary identification. A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, such high level of resolution is likely to introduce heavily imbalanced classes created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. *HiCFeat* did not the address class imbalance. Instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations. <!--This, likely being further exacerbated by the class imbalance.--> Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other re-sampling solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in base pairs between ChIP-seq annotations and genomic bins in TAD boundary prediction, which offer a more spatial measure of association.
